Mastery Learning in Practice:\\ A (Mostly) Descriptive Analysis of Log Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-12-08 paper.tex
Johns comments

Finally, in Section \ref{sec:effects} will will give
quasi-experimental estimates of the effects of reassignment on
students' posttest scores.
We find that reassignment probably decreases student learning,
though the effects vary widely across classrooms.
Section \ref{sec:discussion} will conclude.

\section{The RAND Effectiveness Trial}\label{sec:RANDtrial}
The study to measure the effectiveness of CTAI included 7 states, 73
high schools, and 74 middle schools with nearly 18,700 high school
students and 6,800 middle school students participating. Schools were enrolled in a
total of 52 school districts that were distributed among urban,
suburban, and rural areas. Schools were matched on a set of
covariates, and then randomly assigned to the treatment or control
group. Schools in the control group continued with their current
algebra curriculum, and schools in the treatment group used Carnegie
Learning's curriculum which includes textbook materials as well as
CTAI. Each school participated for two years, with a different cohort
of students participating the second year (with a small fraction of
students present in the study both years because they repeated
algebra). It should be noted that this study did not include statewide
implementations; the study results cannot be generalized to all
schools within the state. In some states, one large school district
participated, while in other states, a set of smaller school districts
participated. The states included Alabama (AL), Connecticut (CT),
Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and
Texas (TX). Each state participated in both the middle school and high
school arms of the study, except AL, which participated only in the
middle school arm. The current study focuses on high school students
only. %\marginpar{I think it reads better to use the spelled-out names,
%  but I'm happy to switch if you feel strongly about this.}%, due to the large number of students in this cohort and the
% significant positive treatment effect previously found for this
% cohort. All discussion that follows refers exclusively to the high
% school data.

There are some limitations to the available data for this study.
Log data from some schools, and some students within schools, were
missing either because the log files were not retrievable, or because
of an imperfect ability to link log data to other study data files.
For this reason, this study uses only data from on the
18 schools for which at least 80\%
students in both study years appear in the log data file.
This sample includes 4460 students, around
100\%
of the total high-school sample.
Table \ref{tab:nByState} gives the number of students in each state of
the sample in each year.
The states in the table are ordered by the total number of students
they represent in the sample; they will appear in this order in
all of the forthcoming tables and figures.
Some figures will only show data from a subset of states; since so few
students were in New Jersey, it will be excluded from almost all
state-by-state comparisons (but included analyses that pool across states).
% latex table generated in R 3.4.2 by xtable 1.8-2 package
% Tue Dec 05 09:39:47 2017
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & TX & KY & MI & LA & CT & NJ \\ 
  \hline
Year 1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
  Year 2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
   \hline
\end{tabular}
\caption{Numbers of students in the sample by state and study year} 
\label{tab:nByState}
\end{table}

In this smaller sample, 164
students do not appear in the log data; they may have not used the CTAI
software at all, or may have been excluded from the log data for other
reasons.
Since we don't know which is true, we exclude these students from most
analyses.

@@

\caption{Percentage of worked problems coming from various levels
  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
  standard and customized variants, denoted by shading.}
\label{fig:curricula}
\end{figure}

@@

Table \ref{tab:medUsage} shows the median numbers of hours,
problems, sections, and units worked on by each student in the dataset
in the two years of the study.
Apparently, usage decreased markedly in the second year: the median
of hours worked decreased by
10
, the median number of problems decreased by
81, and
the median number of sections decreased by
6 from
years 1 to 2.

@@

In contrast to the numbers of hours, problems, and sections students
worked, Table \ref{tab:medUsage} shows that the median number of units
students worked increased by
1
from year 1 to year 2.
This suggests students in year 2 were exposed, on average, to a
slightly wider range of topics.
Figure \ref{fig:unitsByStud} shows boxplots of the numbers of units
worked by state and year.
The geographic variation in units worked mirrors the pattern in Figure
\ref{fig:timeByStud}, with more usage in Kentucky, Michigan, and
Louisiana but less in Texas, Connecticut, and New Jersey.
However, in every state the median Year 2 student worked at least as
many different units as the median Year 1 student.
Variation in the number of units worked also increased slightly
from years 1 to 2---the interquartile range (IQR) increased in every state
except for Kentucky, where a decrease in IQR was accompanied by an
increase in the number of outliers.

@@

Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
the proportions of worked sections in each state and study year that
ended with mastery, promotion, or reassignment, or which were the
student's final section.
In the first year, about
85\% of
worked sections are mastered, except in Connecticut.
Other than in Texas, about
13--15\%
of sections end in promotion.
About 3\% of sections in Texas
and 5\%
in Connecticut end in reassignment, which is even rarer in the other states.

@@

Figure \ref{fig:statusCur} shows the proportions of worked sections
that were mastered or ended in promotion, reassignment, or finality, in
standard and customized versions of each CT curriculum.
Mastery proportions do, indeed, depend on curriculum.
Specifically, students mastered sections from more advanced curricula
less frequently.
Sections from the most basic curriculum, Bridge to Algebra, were
mastered
\% of the time;
those from Algebra I were mastered
83\%
of the time, and those from more advanced curricula were mastered
at a rate of
75\%.
This is unsurprising, since more advanced curricula may be expected
to be more challenging.
However, it may suggest that some students studying advanced topics
would fair better in more standard curricula.

Algebra I sections from customized curricula tended to end in
reassignment more often than sections from the standard Algebra I
curriculum (3\% vs.
1, 1\%).
This may an overall skepticism towards the Carnegie Learning standards
among certain schools and teachers, manifested in both adoption of
alternative curricula and reassignment.%\marginpar{does that make more sense?}

@@

Do teachers reassign students in order to help them catch up with classmates?
According to Figure \ref{fig:classmates}, that might be part of the
story, but isn't all of it.
Across years and states, in about 40\% of
reassignments, at least 75\% of the rest of the class had exited the
section on the same date or earlier.
The figure also reveals that the proportions of students who had graduated or been
promoted from the same section on the same day or earlier was smaller in year 2
than in year 1, especially in Texas and Michigan.

@@

The arrows in the plot represent reassignments.
An arrow from a sending unit to a receiving unit indicates
reassignment from the former to the latter.
The thickness of the arrows represents the proportion of reassignments
originating in the sending unit whose destination was the receiving unit.
The darkness of the arrows represents the number of such reassignments.
For instance, in Figure \ref{fig:trans1}, the arrow from
``4-Quadrant Linear Graphs'' on the left to ``4-Quadrant Linear
Graphs'' on the right is fairly thick, since
44\%
of the reassignments begun in ``4-Quadrant Linear Graphs'' end in the
same unit.
Yet it is also fairly faint, since it only represents
12
reassignments.

Inspecting the figures shows that the most common pattern is for
students to be reassigned to the next unit in the curriculum---this pattern comprises
52\% of the reassignments in year 1 and
24\% of those in
year 2.
On the other hand, it is rare for students to be reassigned within the same unit
(these reassignments account for 15\% and 18\% in the two years).
It is also rare for students to be reassigned to units earlier in the curriculum (comprising
2\% and
8\%).

@@

%\subsection{Average Effects}
Figure \ref{fig:cpYyear} shows students gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
(The number of reassignments was jittered---random noise was added---to avoid overplotting.)
Overall, the relationship between the two variables is positive.
Nevertheless, some non-linearity seems to be present, especially in year 2.
Further, the distribution of the number of reassignments is right-skewed---again, especially in year 2.
Care in modeling the number of reassignments, then, is especially important---observations from students reassigned an unusually large number of times can exert undo influence on a regression model and generate misleading results, particularly in the presence of non-monotonic relationships.
We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for students who were.
Next, $R^{cat}$ defines a categorical variable taking the values $R^{cat}=0,1,2,3,4+$ for students who were reassigned 0, 1, 2, 3 or four or more times, respectively.
Finally, $R^{num}$ is the raw number of reassignments, which we include for completeness.

We used linear models to estimate the effect of reassignment, regressing posttest scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
Since this is unlikely to be the case (even approximately), we ran a
second set of models including student level covariates as well:
pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
15\%). To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2014effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligibility.

@@

But what of unmeasured confounding?
For instance, the negative effect may be due to baseline differences
in ability, beyond what is captured in pretest scores.
\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
Roughly speaking, the idea is to widen the confidence interval from a causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.
As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of posttest scores.
The sensitivity interval for the effect of being reassigned at least
once on posttest scores, allowing for the possible omission of a
hypothetical confounder at most as important as pretest, is
-0.15 $\pm$
0.14.
This interval is quite wide, implying that such a covariate could
explain much of the observed relationship between reassignment and
posttest scores (or that the relationship may be much stronger).
On the other hand, the sensitivity interval allowing for the possible
omission of a less important hypothetical confounder---one that
predicts posttests and reassignment as well as the next best observed
covariates---is
-0.15 $\pm$
0.07.
This interval is substantially tighter.
All in all, unmeasured confounding may play an important role here,
but there is good reason to believe that it does not explain all of
the observed relationship.

The wide variety in the use of reassignment that we have documented
here might suggest that reassignment's treatment effect might vary as
well.
Figure \ref{fig:trtHet} shows estimated classroom-specific treatment
effects of $R^{bin}$, being reassigned at least once.
The estimates came from a multilevel model in which posttest scores
were regressed on $R^{bin}$ and student-level covariates, along with
random effects for classroom and random slopes for $R^{bin}$ varying
by classroom.
Unlike fixed effects models, multilevel models ``partially pool'' data
across classrooms to estimate classroom specific effects more precisely \citep{gelmanHill}.
This is especially important given the small sample sizes within
classrooms.
Figure \ref{fig:trtHet} shows a wide variation in the effect of
reassignment across classrooms---the estimated standard deviation of
these effects was
0.18,
larger than the average effect itself.
While the effect was negative in most classrooms, it was positive in some.
This variation could be due to a number of factors, including
differences in the composition of classrooms, but supports the
hypothesis that differences in when and how reassignment is used
lead to differences in its effect.

@@

The effectiveness of the Cognitive Tutor software presumably depends
on how much, and how, it is used.
This paper exploited available log data from the high-school arm of the CTAI
effectiveness trial---in the second year, an impressive result for the
software---to describe variation and patterns in the software's
usage, paying particular attention to issues of mastery learning.
We found that:
\begin{itemize}
 \item The amount the software was used varied widely between states
   and decreased, overall, from years 1 to 2.
 \item Year 2 saw the proliferation of ``customized'' curricula in
   three states, altering which units students worked, and in what
   order.
 \item Year 2 saw frequent departures from the standard CTAI unit
   sequence, driven mainly, but not entirely, by customized
   curricula.
 \item Section mastery:
 \begin{itemize}
  \item About 85\% of
    worked sections in year 1 and
    81\% in  year 2 ended
    with the student having mastered all of the included skills.
  \item There were three ways students worked a section with
    mastering its contents: exhausting its problems and being
    promoted, being reassigned to a new section by the teacher, and
    ending CT use altogether.
  \item Reassignment was rare, though it was more prevalent in Texas
    and Connecticut than in other states, and more common in year 2
    than 1.
  \item Mastery rates were lower for more advanced curricula.
 \end{itemize}
 \item Reassignment Patterns
  \begin{itemize}
   \item Reassignment rates were determined more by factors varying by state and school
     than by classroom.
   \item Reassignment was more common in the second half of the year
     than in the first---particularly in Texas.
   \item Depending on state and year, reassignment typically takes
     place in one of three scenarios: teachers reassigning students
     who have fallen behind, teachers reassigning (almost) the entire
     class together, and teachers assigning (almost) all students
     \emph{out} of a particular section.
   \item Students are most commonly reassigned to the next unit in the
     curriculum---suggesting that teachers may be advancing lagging
     students---but in some cases they might be finely manipulating
     their students' curricula.
   \end{itemize}
 \item Reassignment appears to lower students' performance on the
   post-test relative to that of their peers; however, the effect
   appears to vary widely between classrooms (and, presumably, how
   reassignment is used).
 \end{itemize}

In practice, mastery learning and topic scaffolding often unfolded
quite differently from the vision of CT's designers.
In some cases, the departures were a matter of a student's inability
to achieve mastery quickly enough, and in other cases departures were
driven by educators' preferences that ran counter to CTAI's design.

>>>>

Finally, in Section \ref{sec:effects} will will give
quasi-experimental estimates of the effects of reassignment on
students' posttest scores.
We find that reassignment probably decreases student learning,
though the effects vary widely across classrooms.
Section \ref{sec:discussion} will conclude the article with a summary
of findings and discussion.

\section{The RAND Effectiveness Trial}\label{sec:RANDtrial}
The study to measure the effectiveness of CTAI included 7 states, 73
high schools, and 74 middle schools with nearly 18,700 high school
students and 6,800 middle school students participating. Schools were enrolled in a
total of 52 school districts that were distributed among urban,
suburban, and rural areas. Schools were matched on a set of
covariates, and then randomly assigned to the treatment or control
group. Schools in the control group continued with their current
algebra curriculum, and schools in the treatment group used Carnegie
Learning's curriculum which includes CTAI textbook materials and sofware. Each school participated for two years, with a different cohort
of students participating the second year (with a small fraction of
students present in the study both years because they repeated
algebra). It should be noted that this study did not include statewide
implementations; the study results cannot be generalized to all
schools within the state. In some states, one large school district
participated, while in other states, a set of smaller school districts
participated. The states included Alabama (AL), Connecticut (CT),
Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and
Texas (TX). Each state participated in both the middle school and high
school arms of the study, except AL, which participated only in the
middle school arm. The current study focuses on high school students
only. %\marginpar{I think it reads better to use the spelled-out names,
%  but I'm happy to switch if you feel strongly about this.}%, due to the large number of students in this cohort and the
% significant positive treatment effect previously found for this
% cohort. All discussion that follows refers exclusively to the high
% school data.

There are some limitations to the available data for this study.
Log data from some schools, and some students within schools, were
missing either because the log files were not retrievable, or because
of an imperfect ability to link log data to other study data files.
For this reason, this study uses only data from the
18 treatment schools for which at least
80\%
of students in both study years appear in the log data file.
This sample includes 4460 students, around
75\%
of the treated high-school sample.
Table \ref{tab:nByState} gives the number of students in the sample by
state and year.
The states in the table are ordered by the total number of students
they represent in the sample; they will appear in this order in
all of the forthcoming tables and figures.
Some figures will only show data from a subset of states; since so few
students were in New Jersey, it will be excluded from almost all
state-by-state comparisons (but included analyses that pool across states).
% latex table generated in R 3.4.2 by xtable 1.8-2 package
% Tue Dec 05 09:39:47 2017
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
 & TX & KY & MI & LA & CT & NJ \\ 
  \hline
Year 1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
  Year 2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
   \hline
\end{tabular}
\caption{Numbers of students in the sample by state and study year} 
\label{tab:nByState}
\end{table}

In this sample from 18 schools, 164
students who participated in the RAND study do not appear in the log data; they may have not used the CTAI
software at all, or may have been excluded from the log data for other
reasons.
Since we don't know which is true, we exclude these students from most
analyses.

@@

\caption{Percentage of worked problems coming from various courses
  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
  standard and customized variants, denoted by shading.}
\label{fig:curricula}
\end{figure}

@@

Table \ref{tab:medUsage} shows the median numbers of hours,
problems, sections, and units worked on by each student in the dataset
in the two years of the study.
Apparently, usage decreased markedly in the second year: the median
of hours worked decreased by
10, the median number of problems decreased by
81, and
the median number of sections decreased by
6.5 from
years 1 to 2.
Yet, as discussed below, the median number of units worked increased
by 1.

@@

In contrast to the decreasing numbers of hours, problems, and sections students
worked in year 2, Table \ref{tab:medUsage} shows that the median number of units
students worked increased by
1
in year 2.
This suggests students in year 2 were exposed, on average, to a
slightly wider range of topics.
Figure \ref{fig:unitsByStud} shows boxplots of the numbers of units
worked by state and year.
The geographic variation in units worked mirrors the pattern in Figure
\ref{fig:timeByStud}, with more usage in Kentucky, Michigan, and
Louisiana but less in Texas, Connecticut, and New Jersey.
However, in every state the median Year 2 student worked at least as
many different units as the median Year 1 student.
Variation in the number of units worked also increased slightly
from years 1 to 2---the interquartile range (IQR) increased in every state
except for Kentucky, where a decrease in IQR was accompanied by an
increase in the number of outliers.

@@

Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
the proportions of worked sections in each state and study year that
ended with mastery, promotion, or reassignment, or as the
student's final section.
In the first year, about
85\% of
worked sections are mastered, except in Connecticut.
Other than in Texas, about
13--15\%
of sections end in promotion.
About 3\% of sections in Texas
and 5\%
in Connecticut end in reassignment, which is even rarer in the other states.

@@

Figure \ref{fig:statusCur} shows the proportions of worked sections
that were mastered or ended in promotion, reassignment, or finality, in
standard and customized versions of each CT curriculum.
Mastery proportions do, indeed, depend on curriculum.
Specifically, students mastered sections from more advanced curricula
less frequently.
Sections from the most basic curriculum, Bridge to Algebra, were
mastered
92\%
of the time;
those from Algebra I were mastered
83\%
of the time, and those from more advanced curricula were mastered
at a rate of
75\%.
This is unsurprising, since more advanced curricula may be expected
to be more challenging.
However, it may suggest that some students studying advanced topics
would fare better in more standard curricula.

Algebra I sections from customized curricula tended to end in
reassignment more often than sections from the standard Algebra I
curriculum (3\% vs.
1.4\%, in year 2).
This may indicate an overall skepticism towards the Carnegie Learning standards
among certain schools and teachers, manifested in both adoption of
alternative curricula and reassignment.%\marginpar{does that make more sense?}

@@

Do teachers reassign students in order to help them catch up with classmates?
According to Figure \ref{fig:classmates}, that might be part of the
story, but isn't all of it.
Across years and states, in about 40\% of
reassignments, at least 75\% of the rest of the class had exited the
section through graduation or promotion on the same date or earlier.
The figure also reveals that the proportions of students who had graduated or been
promoted from the same section on the same day or earlier was smaller in year 2
than in year 1, especially in Texas and Michigan.

@@

The arrows in the plot represent reassignments.
An arrow from a sending unit to a receiving unit indicates
reassignment from the former to the latter.
The thickness of the arrows represents the proportion of reassignments
originating in the sending unit whose destination was the receiving unit.
The darkness of the arrows represents the number of such reassignments.
For instance, in Figure \ref{fig:trans1}, the arrow from
``4-Quadrant Linear Graphs'' on the left to ``4-Quadrant Linear
Graphs'' on the right is fairly thick, since
44\%
of the reassignments from ``4-Quadrant Linear Graphs'' end in the
same unit.
Yet it is also fairly faint, since it only represents
12
reassignments.

Inspecting the figures shows that the most common pattern is for
students to be reassigned to the next unit in the curriculum---this pattern comprises
52\% of the reassignments in year 1 and
24\% of those in
year 2.
On the other hand, it is relatively rare for students to be reassigned within the same unit
(these reassignments account for 15\% and 18\% in the two years).
It is also rare for students to be reassigned to units earlier in the curriculum (comprising
2\% and
8\%).

@@

%\subsection{Average Effects}
Figure \ref{fig:cpYyear} shows students' gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
(The number of reassignments was jittered---random noise was added on
the horizontal dimension---to avoid overplotting.)
Overall, the relationship between the two variables is positive.
Nevertheless, some non-linearity seems to be present, especially in year 2.
Further, the distribution of the number of reassignments is right-skewed---again, especially in year 2.
Care in modeling the number of reassignments, then, is especially important---observations from students reassigned an unusually large number of times can exert undo influence on a regression model and generate misleading results, particularly in the presence of non-monotonic relationships.
We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for students who were.
Next, $R^{cat}$ defines a categorical variable taking the values $R^{cat}=0,1,2,3,4+$ for students who were reassigned 0, 1, 2, 3 or four or more times, respectively.
Finally, $R^{num}$ is the raw number of reassignments, which we include for completeness.

We used linear models to estimate the effect of reassignment, regressing posttest scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
Since this is unlikely to be the case (even approximately), we ran a
second set of models including student level covariates as well:
pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
15\%). To account for
this, we used regression calibration based on the 20 ``multiple
imputations'' used in the original CTAI study,
\citet{pane2014effectiveness}.} race, sex, grade, special education
and gifted status, English as a second language (ESL), and free and reduced-price lunch eligibility.

@@

But what of unmeasured confounding?
For instance, the negative effect may be due to baseline differences
in ability, beyond what is captured in pretest scores.
\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
Roughly speaking, the idea is to widen the confidence interval from an
ostensibly causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.
As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of posttest scores.
The sensitivity interval for the effect of being reassigned at least
once on posttest scores, allowing for the possible omission of a
hypothetical confounder at most as important as pretest, is
-0.15 $\pm$
0.14.
This interval is quite wide, implying that such a covariate could
explain much of the observed relationship between reassignment and
posttest scores (or that the relationship may be much stronger).
On the other hand, the sensitivity interval allowing for the possible
omission of a less important hypothetical confounder---one that
predicts posttests as well as ESL status and reassignment as well as
Hispanic ethnicity, the next best observed
predictors---is
-0.15 $\pm$
0.07.
This interval is substantially tighter.
All in all, unmeasured confounding may play an important role here,
but there is good reason to believe that it does not explain all of
the observed relationship.

The wide variety in the use of reassignment that we have documented
here might suggest that reassignment's treatment effect varies as
well.
Figure \ref{fig:trtHet} shows estimated classroom-specific treatment
effects of $R^{bin}$, being reassigned at least once.
The estimates came from a multilevel model in which posttest scores
were regressed on $R^{bin}$ and student-level covariates, along with
random effects for classroom and random slopes for $R^{bin}$ varying
by classroom.
Unlike fixed effects models, multilevel models ``partially pool'' data
across classrooms to estimate classroom specific effects more precisely \citep{gelmanHill}.
This is especially important given the small sample sizes within
classrooms.
Figure \ref{fig:trtHet} shows a wide variation in the effect of
reassignment across classrooms---the estimated standard deviation of
these effects was
0.18,
larger than the average effect itself.
While the effect was negative in most classrooms, it was positive in some.
This variation could be due to a number of factors, including
differences in the composition of classrooms, but supports the
hypothesis that differences in when and how reassignment is used
lead to differences in its effect.

@@

The effectiveness of the Cognitive Tutor software presumably depends
on how much, and how, it is used.
This paper exploited available log data from the high-school arm of the CTAI
effectiveness trial---which yielded an impressive result for the
software in year 2---to describe variation and patterns in the software's
usage, paying particular attention to issues of mastery learning.
We found that:
\begin{itemize}
 \item The amount the software was used varied widely between states
   and decreased, overall, from years 1 to 2.
 \item Year 2 saw the proliferation of ``customized'' curricula in
   three states, altering which units students worked, and in what
   order.
 \item Year 2 saw frequent departures from the standard CTAI unit
   sequence, driven mainly, but not entirely, by customized
   curricula.
 \item Examination of section mastery found that:
 \begin{itemize}
  \item About 85\% of
    worked sections in year 1 and
    81\% in  year 2 ended
    with the student having mastered all of the included skills.
  \item There were three ways students worked a section with
    mastering its contents: exhausting its problems and being
    promoted, being reassigned to a new section by the teacher, and
    ending CT use altogether.
  \item Reassignment was rare, though it was more prevalent in Texas
    and Connecticut than in other states, and more common in year 2
    than 1.
  \item Mastery rates were lower for more advanced curricula.
 \end{itemize}
 \item Examination of reassignment patterns found that:
  \begin{itemize}
   \item Reassignment rates were determined more by factors varying by state and school
     than by classroom.
   \item Reassignment was more common in the second half of the year
     than in the first---particularly in Texas.
   \item Depending on state and year, reassignment typically takes
     place in one of three scenarios: teachers reassigning students
     who have fallen behind, teachers reassigning (almost) the entire
     class together, and teachers assigning (almost) all students
     \emph{out} of a particular section.
   \item Students are most commonly reassigned to the next unit in the
     curriculum---suggesting that teachers may be advancing lagging
     students---but in some cases they might be finely manipulating
     their students' curricula.
   \end{itemize}
 \item Reassignment appears to lower students' performance on the
   post-test relative to that of their peers; however, the effect
   appears to vary widely between classrooms (and, presumably, how
   reassignment is used).
 \end{itemize}

In practice, mastery learning and topic scaffolding often unfolded
quite differently from the vision of CT's designers.
In some cases, the departures were apparently a matter of a student's inability
to achieve mastery quickly enough, and in other cases of educators' preferences that ran counter to CTAI's design.
